Compare NE-combined vs. NE-separated reference

### 2025-07-09 analyses/cell-type-neuroblastoma-04/exploratory-notebooks/separate-tumor.Rmd
towards prod

In this section, we will compare some of the marker genes identified in the trained SingleR models directly.

When training a model, SingleR performs pairwise comparisons to identify marker genes as the top 10 upregulated genes for each label compared to each other label.

@@

* Marker genes upregulated in NE compared to each other cell type
* Marker genes upregulated in NE-tumor compared to each other cell type

@@

@@

```{r, fig.width = 12}
plot_barcodes <- celltype_df |>
  dplyr::filter(stringr::str_detect(singler_label_separated, "Neuroendocrine")) |>
  dplyr::pull(barcodes)

@@

# Add the separated cell types into colData as `singler_cell_type`
colData(sce) <- colData(sce) |>
  as.data.frame() |>
  dplyr::inner_join(
    celltype_df |>
      dplyr::select(barcodes, singler_cell_type = singler_label_separated),
    by = "barcodes"
  ) |>
  DataFrame(row.names = rownames(colData(sce)))

# plot the NE-only genes
plot1 <- scater::plotExpression(
  sce[different_marker_genes$ne_only, plot_barcodes],
  features = different_marker_genes$ne_only,
  x = "singler_cell_type",
  colour_by = "singler_cell_type",
  point_size = 0.5
) +
  theme(legend.position = "none") +
  ggtitle("Neuroendocrine marker genes")

# plot the NE-tumor--only genes
plot2 <- scater::plotExpression(
  sce[different_marker_genes$ne.tumor_only, plot_barcodes],
  features = different_marker_genes$ne.tumor_only,
  x = "singler_cell_type",
  colour_by = "singler_cell_type",
  point_size = 0.5
) +
  theme(legend.position = "none") +
  ggtitle("Neuroendocrine-tumor marker genes")

@@

Next, we'll look at the NE marker genes between references.

For a quick summary, we'll ask: How many of the NE markers in the combined model (out of 10 total per cell type comparison!) are _also included_ in the separated model's NE (or NE-tumor) marker gene lists?

@@

      # how many combined are in that separated list?
      sum(combined_marker_genes %in% separated_model_markers)
    }
  )
```

Here, we see that NE marker genes from the combined model are essentially entirely contained within the NE and/or NE-tumor categories of the separated model.

@@

This minimal exploration does not suggest that there are salient differences between the two NE categories in the separated model.

@@

In the next section, we'll look at the numbers a bit more closely.

@@

### Score comparisons

In the separated reference, for cells that are labeled either `NE` or `NE-tumor` (considering the `pruned.labels` specifically), which label has the _second highest score_?
We expect that the second highest score will usually be the other `NE` label if these cell types are indeed highly similar.

Note that there were this many total NE and NE-tumor cells identified by the separated model:
```{r}
separated_ne_df <- singler_results_separated |>
  as.data.frame() |>
  tibble::rownames_to_column("barcodes") |>
  # keep only cells labeled with one of the NE categories
  dplyr::filter(stringr::str_detect(pruned.labels, "Neuroendocrine"))
nrow(separated_ne_df)
```

The table below shows counts for how many times the given label is the second-highest score:

```{r}
SingleR::plotDeltaDistribution(
  singler_results_separated,
  show = "delta.next",
  size = 0.5,
  ncol = 4
)
```

```{r}
# tidy the scores into long format
scores_df <- singler_results_separated$scores |>
  as.data.frame() |>
  dplyr::mutate(barcodes = rownames(singler_results_separated)) |>
  tidyr::pivot_longer(
    -barcodes,
    names_to = "score_label",
    values_to = "score"
  )

# This dataframe grabs the labels associated with delta.next, aka the label for the 2nd highest score
next_label_df <- scores_df |>
  dplyr::group_by(barcodes) |>
  # use -score to go in descending order and then keep only the SECOND rank
  dplyr::mutate(score_rank = rank(-score)) |>
  dplyr::filter(score_rank == 2) |>
  dplyr::ungroup() |>
  dplyr::select(barcodes, delta.next_label = score_label)

# This dataframe grabs the actual scores associated with the assigned pruned.labels
full_scores_df <- scores_df |>
  # bring in the real labels and delta.next scores
  dplyr::inner_join(
    data.frame(
      barcodes = rownames(singler_results_separated),
      singler_label = singler_results_separated$pruned.labels,
      delta.next = singler_results_separated$delta.next
    ),
    by = "barcodes"
  ) |>
  # the score whose label matches the assigned label is the score we want
  dplyr::filter(score_label == singler_label) |>
  dplyr::select(barcodes, singler_label_score = score, singler_label, delta.next) |>
  # bring in the delta.next_label
  dplyr::inner_join(
    next_label_df,
    by = "barcodes"
  ) |>
  unique() |>
  # Finally, add a column for the delta.next / singler_label_score
  dplyr::mutate(delta.next_rel_difference = (delta.next / singler_label_score))
full_scores_df |>
  dplyr::filter(delta.next_fraction > 1)

ggplot(full_scores_df) +
  aes(x = singler_label, y = delta.next_fraction) +
  geom_boxplot() +
  theme(axis.text = element_text(angle = 45))
```

```{r}
# [, "Neuroendocrine"]

next_label_df

next_label_barcodes <- next_label_df$barcodes

singler_results_separated |>
  as.data.frame() |>
  tibble::rownames_to_column("barcodes") |>
  dplyr::select(barcodes, pruned.labels, delta.next) |>
  ggplot() +
  aes(x = pruned.labels, y = delta.next) +
  geom_boxplot() +
  theme(axis.text.x = element_text(angle = 90))

dplyr::add_count(delta.next_label)
# format for printing
dplyr::mutate(second_highest_delta_label = stringr::str_remove(cell_type, "scores\\.")) |>
  dplyr::arrange(desc(n)) |>
  dplyr::select(
    assigned_label = pruned.labels,
    second_highest_delta_label,
  )
```

Next we'll look at the score heatmaps to see the overall support for labels as inferred with both models.

#### Separated model SingleR score heatmap

@@

#### Combined model SingleR score heatmap

@@

>>>>

In this section, we will compare some of the marker genes identified in the trained SingleR models directly.
When training a model as we have done here, SingleR performs pairwise comparisons to identify marker genes as the top 10 upregulated genes for each label compared to each other label.

@@

* Marker genes upregulated in `NE` compared to each other cell type
* Marker genes upregulated in `NE-tumor` compared to each other cell type

@@

In the plot, each point represents gene expression in a given cell, and the black diamonds are the mean expression for that gene.

@@

```{r, message=F}
plot_barcodes <- celltype_df |>
  dplyr::filter(stringr::str_detect(singler_label_separated, "Neuroendocrine")) |>
  dplyr::pull(barcodes)

@@

# Add the separated cell types into colData as `singler_cell_type`
colData(sce) <- colData(sce) |>
  as.data.frame() |>
  dplyr::inner_join(
    celltype_df |>
      dplyr::select(barcodes, singler_cell_type = singler_label_separated),
    by = "barcodes"
  ) |>
  DataFrame(row.names = rownames(colData(sce)))
```

```{r, fig.width = 12, message=F, warning=F}
# plot the NE-only genes
plot1 <- scater::plotExpression(
  sce[different_marker_genes$ne_only, plot_barcodes],
  features = different_marker_genes$ne_only,
  x = "singler_cell_type",
  colour_by = "singler_cell_type",
  point_size = 0.5,
) +
  stat_summary(shape = 18) +
  theme(legend.position = "none") +
  ggtitle("Neuroendocrine marker genes")

# plot the NE-tumor--only genes
plot2 <- scater::plotExpression(
  sce[different_marker_genes$ne.tumor_only, plot_barcodes],
  features = different_marker_genes$ne.tumor_only,
  x = "singler_cell_type",
  colour_by = "singler_cell_type",
  point_size = 0.5
) +
  stat_summary(shape = 18) +
  theme(legend.position = "none") +
  ggtitle("Neuroendocrine-tumor marker genes")

@@

Next, we'll look at the NE marker genes between references and ask for a quick summary:

How many of the `NE` markers in the combined model (again, out of 10 total per cell type comparison!) are _also included_ in the separated model's `NE` (or `NE-tumor`) marker gene lists?

@@

      # how many combined are in that separated list?
      sum(combined_marker_genes %in% separated_model_markers)
    }
  ) |>
  as.data.frame() |>
  t()
```

Here, we see that `NE` marker genes from the combined model are essentially entirely contained within the `NE` and/or `NE-tumor` categories of the separated model.

@@

@@

In the subsections below, we'll look at the numbers behind the heatmap a bit more closely.

@@

#### Score heatmaps
Next we'll look at the score heatmaps to see the overall support for labels as inferred with both models.

##### Separated model SingleR score heatmap

@@

##### Combined model SingleR score heatmap

@@

#### Delta median distributions

Finally, we'll look at the distribution of delta.median scores in the separated model specifically for the `NE` categories.

```{r, fig.width = 6}
SingleR::plotDeltaDistribution(
  singler_results_separated,
  labels.use = c("Neuroendocrine", "Neuroendocrine-tumor"),
  size = 1
)
```
